---
title: "ggtraces"
author: "Chenxin Li"
date: '2022-07-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies 
```{r}
library(tidyverse) 
```

This is a tidyverse based workflow 

```{r}
library(RColorBrewer)
```
For colors only, not actually needed. 

# Required input
The workflow requires the input data to be in the tidy format (each row is an observation, and each column is a variable).

It requires at least 3 columns: 

1. The column named `x`, which will be the x axis 
2. the column named `y`, which will be the y axis
3. A `sample` column that indicates the sample ID of each of the traces. 

Addition required values:

1. a vector of sample IDs 
2. x_offset, default = 0.2
3. y_offset, default = 0.4
4. number of traces to plot 

# Functions defined by the workflow
This workflow defines a 6 functions in this order:

1. `find_xy_ranges()` takes the tidy input data frame and finds xmin, xmax, ymin, and ymax. 
2. `make_grid_table()` takes the ranges produced by `find_xy_ranges()` and produce a data frame that will be used to make the coordinate system. Additionally, it requires `x_offset` and `y_offset` and `number_of_traces`. 
3. `make_axis_table()` takes the ranges produced by `find_xy_ranges()` and produce a data frame that will be used to make the coordinate system.
4. `make_coord()` takes the output of `find_xy_ranges()`, `make_grid_table()`, `make_axis_table`, to make a ggplot object that is a blank coordinate system. It also requires `x_offset` and `y_offset` and `number_of_traces`.
5. `map_sample_to_trace()` takes a vector of sample IDs and produce a data frame that maps sample IDs to traces (column of 1 to n). 
6. `plot_traces()` takes the output of all the above and produce a ggplot object. 

# Example data
This section produce two example traces. You can ignore this part. 
```{r}
set.seed(666)

trace1 <- data.frame(
  "x" = seq(0, 2*pi, by = 0.01)
) %>% 
  mutate(y = sin(3*x))

trace2 <- data.frame(
  "x" = seq(0, 2*pi, by = 0.01)
) %>% 
  cbind(
    "amp" = rnorm(n = nrow(trace1), mean = 1, sd = 0.1)
  ) %>% 
  mutate(y = amp*sin(3*x))

example_1 <- rbind(
  trace1 %>% 
    mutate(sample = "sample1"),
  trace2 %>% 
    mutate(sample = "sample2") %>% 
    dplyr::select(-amp)
)

head(example_1)
```

In this example, there are two samples that are just sine waves. 


# Find x and y value ranges 
The following function produces x and y value ranges (min and max). 
```{r}
find_xy_ranges <- function(data){
  data %>% 
    summarise(xmin = min(x),
            xmax = max(x),
            ymin = min(y),
            ymax = max(y))
}
```

```{r}
example_1_ranges <- find_xy_ranges(example_1)

example_1_ranges
```


# Coodinate system 
```{r}
x_offset <- 0.2
y_offset <- 0.4
number_traces <- 3 
```

Note that the example only have 2 traces. 
But we will make space for 3 traces just as example. 

## Make grid and axis tables 
The following function produces a grid 

```{r}
make_grid_table <- function(ranges, x_offset = 0.2, y_offset = 0.4, number_traces){
  data.frame(
     trace = 1:number_traces
  ) %>% 
    mutate(x = ranges$xmin + x_offset*(trace-1)) %>%               
    mutate(xend = ranges$xmax + x_offset*(trace-1)) %>% 
    mutate(y = y_offset*(trace-1)) %>% 
    mutate(yend = y) %>% 
    dplyr::select(-trace) %>% 
    rbind(
      rbind(
      c(ranges$xmin, ranges$xmin + x_offset*(number_traces - 1), 0, y_offset*(number_traces - 1)),
      c(ranges$xmax, ranges$xmax + x_offset*(number_traces - 1), 0, y_offset*(number_traces - 1))
    ) %>% 
      as.data.frame() %>% 
      dplyr::rename(
        x = V1,
        xend = V2,
        y = V3,
        yend = V4
    )
  )
}


```
The following function produce a axis table 

```{r}
make_axis_table <- function(ranges){
  rbind(
    c(ranges$xmin, ceiling(ranges$xmax), -Inf, -Inf), 
    c(-Inf, -Inf, round(ranges$ymin), ranges$ymax)
    ) %>% 
    as.data.frame() %>% 
    dplyr::rename(
      x = V1,
      xend = V2,
      y = V3,
      yend = V4
    )
}
```

```{r}
example1_grid_table <- make_grid_table(example_1_ranges, x_offset = x_offset, y_offset = y_offset, number_traces = number_traces)
example1_axis_table <- make_axis_table(example_1_ranges)
```


The following function produce a ggplot object which is a blank coordinate 
```{r}
make_coord <- function(grid_table, axis_table, ranges, number_traces, x_offset = 0.2, y_offset = 0.4){
  grid_table %>% 
    ggplot(aes(x = x, y = y)) +
    geom_segment(aes(x = x, xend = xend, y = y, yend = yend), 
                 color = "grey75") +
    geom_segment(data = axis_table, 
                  aes(x = x, xend = xend, y = y, yend = yend), 
                  color = "grey25", size = 1.1) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      text = element_text(color = "black", size = 14),
      axis.text = element_text(color = " black"),
      legend.position = "top",
      axis.ticks = element_line(color = "grey20")
  )
}
```

```{r}
example1_coord <- make_coord(
  grid_table = example1_grid_table, 
  axis_table = example1_axis_table,
  ranges = example_1_ranges,
  number_traces = number_traces,
  x_offset = x_offset,
  y_offset = y_offset
)

example1_coord

ggsave("../Results/blank_coord.svg", height = 3, width = 6, bg = "white")
ggsave("../Results/blank_coord.png", height = 3, width = 6, bg = "white")
```

The blank coordinate is a "grob" object. 
We can add ggplot layers to, such as geom, scale, theme, and so on. 

# Produce sample to trace mapping 
Input a vector called `sample_names`. Each element is a trace. 
First element will be the trace in the front. 

```{r}
sample_names <- c("sample1", "sample2")
```

The following function produces sample to trace mapping. 
```{r}
map_sample_to_trace <- function(samples){
  data.frame(
    "sample" = samples
  ) %>% 
    mutate(traces = 1:nrow(.))
}
```

```{r}
example1_mapping <- map_sample_to_trace(sample_names)
example1_mapping
```

# Add traces to blank coordinate 

The following function produces traces given the blank coordinate and everything else. 
```{r}
plot_traces <- function(data, coord, mapping, x_offset = 0.2, y_offset = 0.4, ranges, 
                        x_title = "x value", y_title = "y value", sample_ID_title = "Sample ID"){
  coord +
  geom_line(data = data %>%
              inner_join(mapping, by = "sample") %>%
              mutate(x_adj = x + x_offset*(traces - 1)) %>%
              mutate(y_adj = y + y_offset*(traces- 1)) %>% 
              mutate(sample = reorder(sample, -traces)) %>% 
              mutate(traces = reorder(traces, -traces)), 
            aes(x = x_adj, y = y_adj,
                group = traces, color = sample), 
  size = 0.75) +
  labs(x = x_title,
       y = y_title,
       color = sample_ID_title) +
  scale_color_discrete(limits = mapping$sample)
}
```

```{r}
example_1_traces <- plot_traces(
  data = example_1,
  coord = example1_coord,
  mapping = example1_mapping,
  x_offset = x_offset,
  y_offset = y_offset,
  ranges = example_1_ranges
)

example_1_traces

ggsave("../Results/example_1.svg", height = 3, width = 6, bg = "white")
ggsave("../Results/example_1.png", height = 3, width = 6, bg = "white")
```

This is a grob object. 
We can add more ggplot layers to it if needed, such as replacing the default color palette.  

```{r}
example_1_traces +
  scale_color_manual(values = brewer.pal(8, "Accent")[c(1,5)],
                     limits = example1_mapping$sample) +
  theme(legend.position = "top")

ggsave("../Results/example_1_nicer.svg", height = 3, width = 6, bg = "white")
ggsave("../Results/example_1_nicer.png", height = 3, width = 6, bg = "white")
```


# plot_facet() feature
```{r}
plot_facet <- function(data, x_title = "x", y_title = "y"){
  data %>% 
    ggplot(aes(x = x, y = y)) +
    facet_wrap(~ sample, scales = "free_x", ncol = 1) +
    geom_line(aes(group = sample, color = sample), size = 0.75) +
    labs(x = x_title,
         y = y_title) +
    theme_classic() +
    theme(
        text = element_text(color = "black", size = 14),
        axis.text = element_text(color = " black"),
        legend.position = "none",
        strip.background = element_blank(),
        strip.text = element_text(hjust = 0),
        panel.spacing = unit(0.8, "lines")
    )
} 
```

```{r}
plot_facet(data = example_1) +
  scale_color_manual(values = brewer.pal(8, "Accent")[c(1, 5)])
```
# Pherogram
```{r}
plot_pherogram <- function(data, y_title = "x", legend_title = "y", mapping){
  data %>% 
    mutate(sample = factor(sample, levels = mapping$sample)) %>% 
    ggplot(aes(x = sample, y = x)) +
    geom_tile(aes(fill = y)) +
    scale_fill_gradientn(colors = c(
      "#FFFFFF", "#F0F0F0", "#D9D9D9", "#BDBDBD", "#969696", "#737373", "#525252", "#252525"
   )) +
    scale_y_continuous(trans = "reverse") +
    labs(x = NULL,
        fill = legend_title,
        y = y_title) +
    theme_classic() +
    theme(
        text = element_text(color = "black", size = 14),
        axis.text = element_text(color = " black"),
        legend.position = "right",
        legend.key.height = unit(1.5, "lines") 
        )
}
```

```{r}
plot_pherogram(data = example_1, mapping = example1_mapping)
```



