---
title: "ggtrace_uses"
author: "Chenxin Li"
date: '2022-07-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The easiest way to use ggtrace functions, the best way is running `ggtrace.Rmd` in the same RStudio window in a different tab.
Once the script finishes running, the 6 functions defined by `ggtrace.Rmd` will appear in the environment. 

# Packages
```{r}
library(tidyverse) 
library(patchwork)
library(RColorBrewer)
```

# LC-MS data example
Data from [Li et al., 2022](https://www.biorxiv.org/content/10.1101/2022.07.04.498697v1) 

## Load data
```{r}
LC_MS_data <- read_csv("../Data/Example_data2.csv", col_types = cols()) # col_types = cols() suppresses col type specification message 
head(LC_MS_data)
```

## Rename columns for ggtrace functions  
```{r}
LC_MS_data_2 <- LC_MS_data %>% 
  dplyr::rename(x = `Retention time`,
                y = `Intensity`,
                sample = cell_type)

head(LC_MS_data_2)
```

## Run ggtrace functions one by one
```{r}
example2_ranges <- find_xy_ranges(LC_MS_data_2)

example2_ranges
```

```{r}
example2_grid_table <- make_grid_table(example2_ranges, x_offset = 0.2, y_offset = 0.4, number_traces = 3)
example2_axis_table <- make_axis_table(example2_ranges)

example2_coord <- make_coord(
  grid_table = example2_grid_table, 
  axis_table = example2_axis_table,
  ranges = example2_ranges,
  number_traces = 3,
  x_offset = 0.2,
  y_offset = 0.4
)

example2_coord
```
```{r}
example2_names <- c("Epidermis", "Idioblast")
example2_mapping <- map_sample_to_trace(example2_names)
example2_mapping
```

```{r}
example2_traces <- plot_traces(
  data = LC_MS_data_2,
  coord = example2_coord,
  mapping = example2_mapping,
  x_offset = 0.2,
  y_offset = 0.4,
  ranges = example2_ranges,
  x_title = "Retention time (min)",
  y_title = "Relative\nintensity",
  sample_ID_title = "Base Peak Chromatogram"
)

example2_traces
```

## Final touches 
Manually adjust axis breaks, axis range, color palettes, and axis title position. 
```{r}
example2_traces +
   geom_text(data = example2_mapping, 
            aes(label = sample, 
                x = example2_ranges$xmax + x_offset*(traces-1),
                y = y_offset*(traces-1)),
            size = 4, hjust = -0.1
            ) +
  scale_color_manual(values = brewer.pal(8, "Set2")[c(1,4)],
                     limits = example2_mapping$sample) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  xlim(c(2, 7.5)) +
  theme(legend.position = "none",
        axis.title.y = element_text(hjust = 0.4)) 

ggsave("../Results/LC_MS_example.svg", height = 3, width = 6, bg = "white")
ggsave("../Results/LC_MS_example.png", height = 3, width = 6, bg = "white")
```

Done! 

# Metagene plot example
Data from [Li et al., 2020](https://genome.cshlp.org/content/30/2/173.short) and [Li et al., 2022](https://genome.cshlp.org/content/32/2/309.short)

## Load data 
```{r}
metagene <- read_csv("../Data/metagene.csv", col_types = cols())
head(metagene)
```

## Rename columns for ggtrace
```{r}
metagene_2 <- metagene %>% 
  dplyr::rename(x = `bin start`,
                sample = sample_type) %>% 
  mutate(y = mena_pro_24 * 1000)

head(metagene_2)
```

## Run ggtrace functions one by one
```{r}
example3_ranges <- find_xy_ranges(metagene_2)

example3_ranges
```

```{r}
example3_grid_table <- make_grid_table(example3_ranges, x_offset = 200, y_offset = 150, number_traces = 5)
example3_axis_table <- make_axis_table(example3_ranges)

example3_coord <- make_coord(
  grid_table = example3_grid_table, 
  axis_table = example3_axis_table,
  ranges = example3_ranges,
  number_traces = 5,
  x_offset = 200,
  y_offset = 150
)

example3_coord
```
```{r}
example3_names <- c("sperm", "egg", "zygote", "seedling")
example3_mapping <- map_sample_to_trace(example3_names)
example3_mapping
```

```{r}
example3_traces <- plot_traces(
  data = metagene_2,
  coord = example3_coord,
  mapping = example3_mapping,
  x_offset = 200,
  y_offset = 150,
  ranges = example3_ranges,
  x_title = "Position relative to TSS",
  y_title = "Normalized\ncoverage",
  sample_ID_title = "Cell type"
)

example3_traces
```
## Final touches 
Manually adjust axis breaks, axis range, color palettes, and axis title position. 
```{r}
example3_traces +
  geom_segment(x = -Inf, xend = -Inf, y = 0, yend = 800, size = 1.1, color = "grey20") +
  geom_segment(x = -3000, xend = 2000, y = -Inf, yend = -Inf, size = 1.1, color = "grey20") +
  scale_color_manual(values = c("dodgerblue2", "tomato1", "violetred4", "seagreen"),
                     limits = example3_mapping$sample) +
  scale_y_continuous(breaks = c(0, 200, 400, 600, 800)) +
  theme(legend.position = "top",
        axis.title.y = element_text(hjust = 0.4))

ggsave("../Results/metagene_example.svg", height = 4, width = 5.5, bg = "white")
ggsave("../Results/metagene_example.png", height = 4, width = 5.5, bg = "white")
```

# How does x_offset and y_offset changes the look of the plot?
## Higher x_offset and lower y_offset
```{r}
example2_grid_table_2 <- make_grid_table(example2_ranges, x_offset = 0.4, y_offset = 0.2, number_traces = 3)

example2_coord_2 <- make_coord(
  grid_table = example2_grid_table_2, 
  axis_table = example2_axis_table,
  ranges = example2_ranges,
  number_traces = 3,
  x_offset = 0.4,
  y_offset = 0.2
)

example2_traces_2 <- plot_traces(
  data = LC_MS_data_2,
  coord = example2_coord_2,
  mapping = example2_mapping,
  x_offset = 0.4,
  y_offset = 0.2,
  ranges = example2_ranges,
  x_title = "Retention time (min)",
  y_title = "Relative\nintensity",
  sample_ID_title = "Cell type"
) +
  scale_color_manual(values = brewer.pal(8, "Set2")[c(1,4)],
                     limits = example2_mapping$sample) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  xlim(c(2, 7.5)) +
  theme(legend.position = "top",
        axis.title.y = element_text(hjust = 0.5)) 

example2_traces_2
```
High x_offset and low y_offset facilitate comparisons along y axis. 
It gives the sensation that we are looking at the graph from the side. 

## Higher y_offset than x_offset
```{r}
example2_grid_table_3 <- make_grid_table(example2_ranges, x_offset = 0.1, y_offset = 0.5, number_traces = 3)

example2_coord_3 <- make_coord(
  grid_table = example2_grid_table_3, 
  axis_table = example2_axis_table,
  ranges = example2_ranges,
  number_traces = 3,
  x_offset = 0.1,
  y_offset = 0.5
)

example2_traces_3 <- plot_traces(
  data = LC_MS_data_2,
  coord = example2_coord_3,
  mapping = example2_mapping,
  x_offset = 0.1,
  y_offset = 0.5,
  ranges = example2_ranges,
  x_title = "Retention time (min)",
  y_title = "Relative\nintensity",
  sample_ID_title = "Cell type"
) +
  scale_color_manual(values = brewer.pal(8, "Set2")[c(1,4)],
                     limits = example2_mapping$sample) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  xlim(c(2, 7.5)) +
  theme(legend.position = "top",
        axis.title.y = element_text(hjust = 0.4)) 

example2_traces_3
```
Low x_offset and high y_offset facilitate comparisons along x axis. 
It gives the sensation that we are looking at the graph from the top. 

## side by side comparison
```{r}
patchwork::wrap_plots(
  example2_traces_2, example2_traces_3, nrow = 2 
)  

ggsave("../Results/LC_MS_perspectives.svg", height = 6, width = 6, bg = "white")
ggsave("../Results/LC_MS_perspectives.png", height = 6, width = 6, bg = "white")
```



# Facet plot
```{r}
plot_facet(LC_MS_data_2, x_title = "Retention time (min)", y_title = "Relative intensity") +
  scale_color_manual(values = brewer.pal(8, "Set2")[c(1,4)]) 

ggsave("../Results/LC_MS_facet.svg", height = 3.5, width = 5, bg = "white")
ggsave("../Results/LC_MS_facet.png", height = 3.5, width = 5, bg = "white")
```
# Pherogram
 
```{r}
plot_pherogram(data = metagene_2, 
               y_title = "Position relative to TSS", 
               legend_title = "Normalized\ncoverage",
               mapping = example3_mapping)

ggsave("../Results/metagene_pherogram.svg", height = 5, width = 4.75, bg = "white")
ggsave("../Results/metagene_pherogram.png", height = 5, width = 4.75, bg = "white")
```


